\documentname: Is machine learning good or bad for the natural sciences?

### 2023-12-26 paper/main.tex
tweaks to the appendix and etc

\begin{abstract}
  Machine learning (ML) methods are having a huge impact across all of the sciences.
  However, ML has a strong ontology---in which only the data exist---and a strong epistemology---in which a model is considered good if it performs well on held-out training data.
  These philosophies are in strong conflict with both standard practices and key philosophies in the natural sciences.
  Here we identify locations for ML in the natural sciences where the ontology and epistemology are valuable.
  For example, when an expressive machine-learning model is used in a causal inference to represent the effects of confounders, such as foregrounds, backgrounds, or instrument calibration parameters, the model capacity and loose philosophy of ML can sometimes make the results more trustworthy.
  We also show that the there are contexts in which the introduction of ML introduces strong, unwanted statistical biases.
  For one, when expressive models are used to emulate physical (or first-principles) simulations, they introduce strong confirmation biases.
  For another, when expressive regressions are used to label datasets, those labels cannot be used in downstream joint analyses without taking on uncontrolled biases.
  These remarks are intended to apply to all of the natural sciences, but most of the specific examples are in the astrophysics domain.
\end{abstract}

@@

\paragraph{Acknowledgements:}
It is a pleasure to thank
  Gaby Contardo (SISSA),
  Jennifer Hill (NYU),
  Adrian Price-Whelan (Flatiron),
  Sam Roweis (deceased),
  Bernhard Sch\"olkopf (MPI-IS), and
  Soledad Villar (JHU)
for conversations over the years related to these arguments.

@@

\clearpage\appendix
\section{Population-level biases in regression outputs}\label{app:toy}
A machine-learning (ML) regression learns the best set of parameters (weights and thresholds, for example) $\hat{\theta}$ of an expressive function $f(x;\theta)$ from a training set of labeled data.
The learned function $f(x;\hat{\theta})$ delivers a predicted or estimated label $\hat{y}$ for any data point with a complete set of features $x$.
Thus an estimated label $\hat{y}_\ast$ for a data point with features $x_\ast$ has taken information \emph{both} from the features $x_\ast$ and \emph{also} from the features and labels of all the data points in the training set used to set the parameters $\hat{\theta}$.
In applications in which regression-estimated labels are used, it matters \emph{how much information} is coming from the individual data-point's features and how much from the original training set.

@@

>>>>

\begin{abstract}
  Machine learning (ML) methods are having a huge impact across all of the sciences.
  However, ML has a strong ontology---in which only the data exist---and a strong epistemology---in which a model is considered good if it performs well on held-out training data.
  These philosophies are in strong conflict with both standard practices and key philosophies in the natural sciences.
  Here we identify locations for ML in the natural sciences where the ontology and epistemology are valuable.
  For example, when an expressive machine-learning model is used in a causal inference to represent the effects of confounders, such as foregrounds, backgrounds, or instrument calibration parameters, the model capacity and loose philosophy of ML can sometimes make the results more trustworthy.
  We also show that the there are contexts in which the introduction of ML introduces strong, unwanted statistical biases.
  For one, when ML models are used to emulate physical (or first-principles) simulations, they introduce strong confirmation biases.
  For another, when expressive regressions are used to label datasets, those labels cannot be used in downstream joint analyses without taking on uncontrolled biases.
  These remarks are intended to apply to all of the natural sciences, but most of the specific examples are in the astrophysics domain.
\end{abstract}

@@

\paragraph{Acknowledgements:}
It is a pleasure to thank
  Gaby Contardo (SISSA),
  Jennifer Hill (NYU),
  Adrian Price-Whelan (Flatiron),
  Hans-Walter Rix (MPIA),
  Sam Roweis (deceased),
  Bernhard Sch\"olkopf (MPI-IS), and
  Soledad Villar (JHU)
for conversations over the years related to these arguments, and
  Vedant Chandra (Harvard)
for discussions of the bias in regression outputs.

@@

\clearpage\appendix
\section{Population-level biases in regression outputs}\label{app:toy}
A machine-learning (ML) regression learns the best (that it can find) value of a list or vector $\hat{\theta}$ of parameters (weights and thresholds, for example) of an expressive function $f(x;\theta)$ that takes as input a feature list or vector $x$ and outputs a label estimate $y$.
The parameters $\hat{\theta}$ are learned from a training set of labeled data, which is a set of $(x, y)$ pairs.
The learned function $f(x;\hat{\theta})$ delivers a predicted or estimated label $\hat{y}$ for any data point with a complete set of features $x$.
Thus a particular estimated label $\hat{y}_\ast$ for a particular data point with features $x_\ast$ obtains information \emph{both} from the features $x_\ast$ and \emph{also} from the features and labels of all the data points in the training set used to set the parameters $\hat{\theta}$.
In applications in which regression-estimated labels are used, it matters \emph{how much information} is coming from the individual data-point's features and how much from the original training set.

@@

HOGG: COMMENT OR SHOW that the bias goes down as the information in the features goes up (relative to the information in the training-set distribution (the prior)).
